Leadership across cultures and contexts: an exploratory analysis

### 2020-05-04 leadership_across_cultures_contexts.Rmd
Minor change in discussion.

---
title: "Leadership across cultures and contexts: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: Leader2.bib
csl: evolution-and-human-behavior.csl
---

@@

Our second goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing proportions and 95% CIs as above and to estimate how costs and benefits varied by group structure types.

Our third goal was to assess the extent to which each of the leadership trait and cost/benefit variables varied by four theoretically important factors for which we had data for all rows: subsistence strategy, continental region, group context, and leader sex.

Our fourth goal was to extract higher-order dimensions of leader qualities and functions from our researcher coded variables. We used several methods to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

Our fifth goal was to assess variation in our "feature" dimensions by subsistence strategy, continental region, group context, and leader sex.

@@

To supplement results from the aforementioned methods, we also performed textual analyses and explored variation measures of biases stemming from the ethnographic record that may have impacted results.

@@

The analyses computing levels of evidence for each variable used the entire set of text records. Analyses investigating lower dimensional representations used a smaller set that omitted rows with no evidence for any variable (i.e., all zeros), and variables with very little evidence (almost all zeros). All analyses were conducted with `r R.version.string`.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using intercept-only logistic mixed effects models that included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

```{r leadertotalbenefits, fig.width=10, fig.cap="Estimated marginal means of a logistic mixed effects regression model of leader total benefits as a function of A. subsistence strategy, B. group structure, C. continental region, and D. leader sex. Model included random intercepts for ethnographer nested within culture. Values are on the response scale."}
# p1 <- 
#   hagenutils::ggemmeans(emmeans(m_ldrtotben, type = 'response', specs = 'group.structure2')) +
#   scale_x_continuous(limits = c(0, 0.1)) +
#   labs(title = "Total leader benefits", x = "", y = "")
# 
# p2 <- hagenutils::ggemmeans(emmeans(m_ldrtotcost, type = 'response', specs = 'group.structure2')) +
#   scale_x_continuous(limits = c(0, 0.1)) +
#   labs(title = "Total leader costs", x = "\nProbability", y = "")
# 
# p1 + p2 + plot_layout(ncol = 1) + plot_annotation(tag_levels = 'A')

@@

```{r leadertotalcosts, fig.cap="Estimated marginal means of a logistic mixed effects regression model of leader total costs as a function of A. subsistence strategy, and B. group structure. Model included random intercepts for ethnographer nested within culture. Values are on the response scale."}
theplots <- all_emms(m_ldrtotcost, c('subsistence', 'group.structure2'), 0.06, 'Predictors of leader costs')

@@

```{r groupstatus, fig.width=10, fig.cap="Estimated marginal means of a mixed effects logistic regression model of evidence for high status (0/1) as a function of A. subsistence strategy, B. group structure, C. continental region, and D. leader sex. Model included random intercepts for ethnographer nested within culture. Values are on the response scale (probability)."}
# ((plot_emm_status_group2 / plot_emm_status_sex2) | (plot_emm_status_subsistence2 / plot_emm_status_region2)) + plot_annotation(title = "Predictors of high status", tag_levels = 'A')

@@

Although only a small fraction of the ethnographic record sampled here discussed female leaders, our results do provide some insight on the nature of female leadership across human societies and sex-differences in leadership.

Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families by @garfield_evolutionary_2019 and @hagen_leadership_2019. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

@@

```{r heatmapregion, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by each continental region. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r heatmapsex, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by leader sex. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r heatmapgroups, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by each group context. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r multiout}
multi_out <-
  multi_allvars %>% 
  select(
    var,
    Variable, 
    Evidence, 
    interceptAIC, 
    multiAIC, 
    AIC_diff,
    pvalue_subsis,
    pvalue_region,
    pvalue_sex,
    pvalue_groups
    ) %>% 
  mutate(
    var = map_chr(var, ~ str_split(., '_')[[1]][[1]])
  ) %>% 
  arrange(Variable)
```

```{r variablevars}
multi_out %>% 
  dplyr::filter(AIC_diff < -2) %>% 
  kable(
  caption = 'Retained logistic mixed effects models of each variable as functions of subsistence strategy, continental region, leader sex, and group structures. Models were retained if their AIC values decreased by two or more relative to an intercept-only model. Evidence: the number of rows with evidence for the variable. interceptAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each predictor variable in the model.',
  digits = 2
  )
```

```{r universalvars}
multi_out %>%
  dplyr::filter(AIC_diff >= -2) %>% 
  kable(
  caption = 'Logistic mixed effects models that did not show improvement in AIC values relative to an intercept-only model. These study variables either did not vary much by subsistence strategy, continental region, leader sex, or group structures, or had little evidence. Evidence: the number of rows with evidence for the variable. interceptAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each predictor variable in the model.',
  digits = 2
  )

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Mediate, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/material success. Prosocial competencies was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).

>>>>

---
title: "Leadership across cultures and contexts: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2
bibliography: Leader2.bib
csl: evolution-and-human-behavior.csl
---

@@

Our second goal was to extract higher-order dimensions of leader qualities and functions from our researcher coded variables. We used several methods to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

With an interpretable method of dimensionality reduction, we aimed to predict variation in our "feature" dimensions from population-level variation, such as subsistence type, region, and group structure.

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing proportions and 95% CIs as above and to estimate how costs and benefits varied by group structure types and other population-level measures.

To supplement results from the aforementioned methods, we also performed textual analyses and explored variation in measures of biases stemming from the ethnographic record that may have impacted results.

@@

The analyses computing levels of evidence for each variable used the entire set of text records. Analyses investigating lower dimensional representations used a slightly smaller set that omitted records with no evidence for any variable (i.e., all zeros). All analyses were conducted with `r R.version.string`.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- publication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

```{r leadertotalbenefits, fig.width=10, fig.cap="Estimated marginal means of A. leader total benefits, and B. total costs, as a function of group structure, based on a logistic mixed effects regression model with random intercepts for ethnographer nested within culture. Values are on the response scale."}
# p1 <- 
#   hagenutils::ggemmeans(emmeans(m_ldrtotben, type = 'response', specs = 'group.structure2')) +
#   scale_x_continuous(limits = c(0, 0.1)) +
#   labs(title = "Total leader benefits", x = "", y = "")
# 
# p2 <- hagenutils::ggemmeans(emmeans(m_ldrtotcost, type = 'response', specs = 'group.structure2')) +
#   scale_x_continuous(limits = c(0, 0.1)) +
#   labs(title = "Total leader costs", x = "\nProbability", y = "")
# 
# p1 + p2 + plot_layout(ncol = 1) + plot_annotation(tag_levels = 'A')

@@

```{r leadertotalcosts}
theplots <- all_emms(m_ldrtotcost, c('subsistence', 'group.structure2'), 0.06, 'Predictors of leader costs')

@@

```{r groupstatus, fig.width=10, fig.cap="Estimated marginal means of a mixed effects logistic regression model of evidence for high status (0/1) as a function of group type, subsistence strategy, continental region, and leader sex, with random effects for ethnographer nested within culture. Values are on the response scale (probability)."}
# ((plot_emm_status_group2 / plot_emm_status_sex2) | (plot_emm_status_subsistence2 / plot_emm_status_region2)) + plot_annotation(title = "Predictors of high status", tag_levels = 'A')

@@

Only a small fraction of the ethnographic record sampled here discussed female leaders. Our results do however provide some insight on female leadership and sex-differences in leadership across cultures and contexts. Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families by @garfield_evolutionary_2019 and @hagen_leadership_2019 and previous cross-cultural studies [e.g., @yanca_female_2004]. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

@@

```{r heatmapregion, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by each subsistence strategy. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r heatmapsex, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by each subsistence strategy. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r heatmapgroups, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each outcome variable by each subsistence strategy. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}

@@

```{r variablevars}

multi_aic %>% 
  select(
    var,
    Variable, 
    Sum, 
    uniAIC, 
    multiAIC, 
    AIC_diff,
    pvalue_subsis,
    pvalue_region,
    pvalue_female,
    pvalue_groups
    ) %>% 
  mutate(
    var = map_chr(var, ~ str_split(., '_')[[1]][[1]])
  ) %>% 
  arrange(Variable) %>% 
  kable(
  caption = 'Retained logistic mixed effects models of each variable as functions of subsistence strategy, continental region, leader sex, and group structures. Models were retained due to improvement in AIC values relative to an intercept-only model. Sum: the number of rows with evidence for the variable. uniAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each explanatory variable in the model.',
  digits = 2
  )
```

```{r universalvars}
multi_allvars %>%
  dplyr::filter(AIC_diff >= -2) %>% 
  select(
    var,
    Variable, 
    Sum, 
    uniAIC, 
    multiAIC, 
    AIC_diff,
    pvalue_subsis,
    pvalue_region,
    pvalue_female,
    pvalue_groups
    ) %>% 
  mutate(
    var = map_chr(var, ~ str_split(., '_')[[1]][[1]])
  ) %>% 
  arrange(Variable) %>% 
  kable(
  caption = 'Logistic mixed effects models that did not show improvement in AIC values relative to an intercept-only model. These study variables either did not vary much by subsistence strategy, continental region, leader sex, or group structures, or had little evidence. Sum: the number of rows with evidence for the variable. uniAIC: AIC of intercept-only model. multiAIC: AIC of multivariate model. AIC_diff: difference in AIC values. P-values are the unadjusted p-values of each explanatory variable in the model.',
  digits = 2
  )

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. The Mediate, Prosociality, and Organize features were modestly positively correlated with each other, but not with Prestige or Social/material success. Big man was correlated with Cultural conformity. See Figure \@ref(fig:featurecor).
